---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

Load the needed R libraries:

```{r}
library(dplyr)
library(ggplot2)
library(scales)
```

Here is the URL for the source data .zip archive:

```{r}
sourceDataZipURL <- 
    "https://d396qusza40orc.cloudfront.net/repdata/data/activity.zip"
```

Download the .zip archive if it's not alreadly available locally.

```{r}
zipFile <- basename(sourceDataZipURL)
if (!file.exists(zipFile)) {
    download.file(sourceDataZipURL, zipFile, method = "curl")
}
```

Name of the file to extract from the .zip archive.

```{r}
sourceDataFile <- "activity.csv"
```

Extract the data file from the .zip archive, and load the data, using
read.csv, from the data file into a dplyr data frame tbl.

```{r}
data_1 <- tbl_df(read.csv(unz(zipFile, sourceDataFile)))
str(data_1)
```

Let's convert the date variable, currently of class factor, into class date.

```{r}
data_1$date <- as.Date(data_1$date)
```

And lets add a new variable that identifies the day of week for each date.

```{r}
data_1 <- data_1 %>%
          mutate("day_of_week" = as.factor(weekdays(date)))

str(data_1)
```

Now the dataset is ready to be explored. 

## What is mean total number of steps taken per day?

In order to ignore the NAs for this part of the assignment, it would be
helpful to know where they are.  We'll look for NAs in each of the steps,
date, and interval variables.

```{r}
any(is.na(data_1$steps))
any(is.na(data_1$date))
any(is.na(data_1$interval))
```

Only the steps variable contains NAs.

Now let's see if the observations associated with the dates containing NAs 
are all NAs or if they have some non-missing step counts.

```{r}
data_1 %>%
filter(is.na(steps)) %>%
group_by(date) %>%
summarize(total_NAs = sum(is.na(steps)))
```

We see that there are eight dates with 288 missing step counts each.

Now we'll see how many total intervals are associated with each date.

```{r}
interval_counts_by_date <- data_1 %>%
                           group_by(date) %>%
                           summarize(interval_count = sum(!is.na(interval)))

unique(interval_counts_by_date$interval_count) 
                                              
```

Using "!is.na(interval)" in the dplyr command chain above is a simple way of
counting the number of intervals for each date, since we already know there
are no NA values in the interval column.  Since the unique command returned
only one value, every date has exactly 288 intervals.  That means that for
dates with missing step counts, *all* step counts are missing.

So now that we know where the NAs are in the data,  we'll pick a reasonable
technique for ignoring them.  Using a chain of dplyr commands, we'll find the
total number of steps for each date. We'll let the NAs propagate in the
summarize command.  If we chose to set na.rm=TRUE in the summarize command,
we'd be setting the total_steps value to zero rather than NA for those dates,
which would artifically lower the mean and median calculations in the next
steps.  I think that would incorrectly introduce bias into the mean and
median counts of total steps per day, since they are missing and we don't
know that they would be zero.  When we  make the histogram, ggplot will by
default ignore the propagated NA values.

```{r}
total_steps_by_date_1 <- data_1 %>%
                         group_by(date) %>%
                         summarize(total_steps = sum(steps, na.rm = FALSE))
```


Make a histogram of the total number of steps taken each day.

```{r plot1, fig.height=4}
binwidth_val = round(max(total_steps_by_date_1$total_steps, na.rm=TRUE)/30)
title_text <- "Histogram of Total Steps Taken Each Day"
title_text <- paste(title_text, "\n(Missing Values Ignored)")

g <- ggplot(total_steps_by_date_1, aes(x=total_steps, fill="salmon"))
g <- g + geom_histogram(binwidth=binwidth_val)
g <- g + labs(x = "Total Steps", y = "Count")
g <- g + labs(title = title_text)
g <- g + theme(legend.position="none")
plot(g)
```

Now we'll calculate the mean and median total number of steps per day.  We'll
ignore the eight dates with total_steps values that are set to NA.

```{r}
mean_total_steps_by_day_1 <- mean(total_steps_by_date_1$total_steps, 
                                  na.rm = TRUE)
print(mean_total_steps_by_day_1)

median_total_steps_by_day_1 <- median(total_steps_by_date_1$total_steps, 
                                      na.rm = TRUE)
print(median_total_steps_by_day_1)
```

The **mean** total number of steps per day is 
**`r format(mean_total_steps_by_day_1, scientific = FALSE)`**.

The **median** total number of steps per day is 
**`r format(median_total_steps_by_day_1, scientific = FALSE)`**.


## What is the average daily activity pattern?

We'll make a time series plot of the five-minute interval (x-axis) and the
average number of steps taken, averaged across all days (y-axis).

To do this, we'll first calculate the average number of steps for each
interval across all dates, sorted in descending order of average step count.

```{r}
start_time <- strptime("00:00", "%H:%M")
mean_steps_by_interval <- data_1 %>%
                          group_by(interval) %>%
                          summarize(mean_steps = mean(steps, na.rm=TRUE)) %>%
                          arrange(interval) %>%
                          mutate("time_of_day" = seq(start_time, 
                                                     length.out=288, 
                                                     by="5 mins")) %>%
                          arrange(desc(mean_steps))
```

**Note that intervals on the plot's x-axis will be mapped to time of day.**

The intervals as brought in from the source data set progress like this:

0, 5, 10, ... , 50, 55, 100, 105, ... , 150, 155, 200, ... , 2350, 2355

There is a discontinuous jump from **55 to **00 every time the two
least-significant digits of the interval reach 55 minutes.  We could just
plot the step counts againt the raw interval values on the x-axis (and Dr.
Peng's example panel plot appears to do just that with an x-axis range from 0
to 2355), but I am going to choose to convert the interval format into a
POSIXct time of day variable.  By doing this, datapoints will be plotted at
consistent intevals on the x-axis as time progresses, and a bit of
distoration from the **55 to **00 jumps will be avoided.  This is a minor
issue, but I think that mapping the interval number to time of day will
result in a time series plot that is slightly more faithful to the source
data.

You can see that time_of_day is a new variable added to my data frame above.

Now the code for building the plot:

```{r plot2, fig.height=4}
title_text <- "Time Series at Five-minute Intervals Showing Number"
title_text <- paste(title_text, "\nof Steps Taken Averaged Across All Days")

g <- ggplot(mean_steps_by_interval, aes(x=time_of_day, y = mean_steps))
g <- g + geom_line(color = "blue")
g <- g + labs(x = "Time of Day",
              y = "Average Steps Taken")
g <- g + labs(title = title_text)
g <- g + scale_x_datetime(labels = date_format("%H:%M"), 
                          breaks = "4 hours",
                          minor_breaks = "1 hours"
                         )
plot(g)
```

The interval with the maximum average number of steps is at the top of
the sorted data frame.

```{r}
interval_with_max_number_of_average_steps <- mean_steps_by_interval$interval[1]
print(interval_with_max_number_of_average_steps)

max_number_of_average_steps <- mean_steps_by_interval$mean_steps[1]
print(max_number_of_average_steps)
```

So the 5-minute interval, on average across all the days in the
dataset, which contains the maximum number of steps is interval 
number **`r interval_with_max_number_of_average_steps`** with 
**`r round(max_number_of_average_steps)`** steps.  Interval 835 corresponds
to 8:35 a.m.

## Imputing missing values

Find the total number of rows with missing values.

```{r}
total_rows_with_missing_values <- sum(is.na(data_1))
print(total_rows_with_missing_values)
```

The total number of rows with missing values is 
**`r total_rows_with_missing_values`**.

**Strategy for the imputation of missing data:**

Recall that we showed that each date that is missing step counts is actually
missing all the counts for that day.  So it seems reasonable to replace the
missing counts with the averages for that day.  We'll create a new data set
where for every interval that is missing a step count, we'll replace the
missing value with the mean for that day of the week.

```{r}
mean_steps_by_day_of_week <- data_1 %>%
                             group_by(day_of_week) %>%
                             summarize(mean_steps = mean(steps, na.rm=TRUE))

data_2 <- data_1
for (i in 1:nrow(data_1)) {
    if (is.na(data_2$steps[i])) {
        data_2$steps[i] <- (
            mean_steps_by_day_of_week$mean_steps[
                mean_steps_by_day_of_week$day_of_week == data_2$day_of_week[i]
            ]
        )
    }
}
```

With the missing step counts replaced, we'll make a histogram of the total
number of steps taken each day for the case with imputed missing values.

```{r plot3, fig.height=4}

total_steps_by_date_2 <- data_2 %>%
                         group_by(date) %>%
                         summarize(total_steps = sum(steps))

binwidth_val = round(max(total_steps_by_date_2$total_steps)/30)
title_text <- "Histogram of Total Steps Taken Each Day"
title_text <- paste(title_text, "\n(Missing Values Imputed)")

g <- ggplot(total_steps_by_date_2, aes(x=total_steps, fill="salmon"))
g <- g + geom_histogram(binwidth=binwidth_val)
g <- g + labs(x = "Total Steps", y = "Count")
g <- g + labs(title = title_text)
g <- g + theme(legend.position="none")
plot(g)
```

With the imputed values in place, we'll now find the mean and median total 
number of steps per day and the change from the estimates without the imputed
values.

```{r}
mean_total_steps_by_day_2 <- mean(total_steps_by_date_2$total_steps)
print(mean_total_steps_by_day_2)

mean_total_steps_by_day_percent_change <-
    ((mean_total_steps_by_day_2 - mean_total_steps_by_day_1)/
     mean_total_steps_by_day_1) * 100
print(mean_total_steps_by_day_percent_change)

median_total_steps_by_day_2 <- median(total_steps_by_date_2$total_steps)
print(median_total_steps_by_day_2)

median_total_steps_by_day_percent_change <- 
    ((median_total_steps_by_day_2 - median_total_steps_by_day_1) / 
     median_total_steps_by_day_1) * 100 
print(median_total_steps_by_day_percent_change)
```

With the imputed values, the **mean** total number of steps per day is now
**`r format(mean_total_steps_by_day_2, scientific = FALSE)`**, 
which is a change of 
**`r sprintf("%0.2f", mean_total_steps_by_day_percent_change)`%**
compared with the previous value of
**`r format(mean_total_steps_by_day_1, scientific = FALSE)`**.

With the imputed values, the **median** total number of steps per day is now 
**`r format(median_total_steps_by_day_2, scientific = FALSE)`**,
which is a change of 
**`r sprintf("%0.2f", median_total_steps_by_day_percent_change)`%**
compared with the previous value of
**`r format(median_total_steps_by_day_1, scientific = FALSE)`**.

## Are there differences in activity patterns between weekdays and weekends?

We'll create a new factor variable in the dataset with two 
levels -- "weekday" and "weekend".

```{r}
data_2 <- data_2 %>%
          mutate("day_type" = 
              ifelse(
                  ((day_of_week == "Sunday") | (day_of_week == "Saturday")),
                  "weekend",
                  "weekday"
              )
          )
data_2$day_type = as.factor(data_2$day_type)
```

Reverse the default factor order for day_type, putting "weekend" before
"weekday".

```{r}
new_order = c("weekend", "weekday")
data_2$day_type <- 
    factor(data_2$day_type, levels = new_order)

str(data_2)
```

Now we'll find the average number of steps for each interval across all 
weekdays and across all weekend days .

```{r}
start_time <- strptime("00:00", "%H:%M")
mean_steps_by_interval_by_day_type <- data_2 %>%
                                      group_by(day_type, interval) %>%
                                      summarize(mean_steps = mean(steps)) %>%
                                      arrange(interval) %>%
                                      mutate("time_of_day" = seq(start_time, 
                                                                 length.out=288,
                                                                 by="5 mins"
                                                                )
                                      )
```

Make a panel plot containing a time series plot of the 5-minute interval
(x-axis) and the average number of steps taken, averaged across all weekday
days or weekend days (y-axis).

**Note that as with the first time series plot, intervals on the
x-axis are mapped to time of day.**

```{r plot4}
title_text <- "Time Series at Five-minute Intervals Showing Average"
title_text <- paste(title_text, "\nNumber of Steps Taken on")
title_text <- paste(title_text, " Weekdays and Weekends")

g <- ggplot(mean_steps_by_interval_by_day_type, aes(x = time_of_day, 
                                                    y = mean_steps))
g <- g + geom_line(color = "blue")
g <- g + facet_wrap(~ day_type, nrow = 2, ncol = 1)
g <- g + labs(x = "Time of Day", y = "Average Steps Taken")
g <- g + labs(title = title_text)
g <- g + scale_x_datetime(labels = date_format("%H:%M"), 
                          breaks = "4 hours",
                          minor_breaks = "1 hours"
                         )
plot(g)
```
